---
title: "Algorithms for Quantitative Pedology"
author: "D.E. Beaudette, J. Skovlin, A.G. Brown, P. Roudier, and S.M. Roecker"
output:
  bookdown::html_document2:
    fig_caption: yes
    number_sections: no
    theme: journal
  pdf_document:
    fig_caption: yes
  html_document:
    df_print: paged
  word_document:
    fig_caption: yes
bibliography: bibliography.bib
---

```{r include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE, 
  warning = FALSE, 
  background = '#F7F7F7', 
  fig.align = 'center', 
  dev = 'svglite', 
  fig.ext = ".svg", 
  tidy = FALSE, 
  verbose = FALSE
)
```

```{r echo = FALSE, eval = FALSE}
# re-make all figures
source('figs-SPC.R')
source('figs-soil-morphology.R')
source('figs-change-in-support.R')
source('figs-sketches-num-taxonomy.R')


# convert figures to PNG for final export to Word
source('convert-figs-to-PNG.R')
```


<!-- ~ 140 words -->
## Abstract

The Algorithms for Quantitative Pedology (AQP) project is a suite of packages for the R programming language that simplify quantitative analysis of soil profile data. The "aqp" package provides a vocabulary (functions and data structures) tailored to the complexity of soil profile information. The "soilDB" package provides interfaces to databases and web services; leveraging the "aqp" vocabulary. The "sharpshootR" package provides tools to assist with summary and visualization. Bridging the gap between pedometric theory and practice is central to the purpose of the AQP project. The AQP R packages have been extensively tested and documented, applied to projects involving hundreds of thousands of soil profiles, and integrated into widely-used tools such as SoilWeb. These packages serve an important role in routine data analysis within the U.S. Department of Agriculture and in other soil survey programs worldwide.


```{r child = 'introduction.Rmd'}
```

<!-- this now includes change of support -->
```{r child = 'SPC-objects.Rmd'}
```


```{r child = 'soil-morphology.Rmd'}
```


```{r child = 'numerical-classification.Rmd'}
```


```{r child = 'water-balance.Rmd'}
```


```{r child = 'conclusions.Rmd'}
```


# References

