
<!-- ~2000 words -->
## Soil Morphology

The field description of a soil profile (genetic horizon depths, boundaries, color, soil texture, rock fragment volume, structure, etc.) is typically the foundation upon which additional sampling, laboratory characterization, or soil survey are based. In aggregate, a complete collection of horizons, associated properties, and landscape context (e.g. catenary position or other geomorphic description) represent an atomic unit of pedologic inquiry: the pedon [@SoilSurveyManual2017]. The AQP family of R packages and the `SoilProfileCollection` datatype were designed specifically to elevate the pedon (and collections of pedons) to a convenient abstraction (an object), enabling a simpler interface to what would otherwise be a complex hierarchy of above and below-ground records. In *aqp*, The more generic term "profile" is used instead of pedon to accommodate incomplete data (missing above-ground information) or otherwise truncated (e.g. 0-25cm) observations. Central to this approach are the specification of profile IDs and horizon depths, above-ground ("site") vs. below-ground ("horizon") attributes, and ideally horizon designation and soil color.



### Soil Color
The color of soil material observed during field investigations is one of the most striking and useful properties recorded as part of a soil profile description. Typically recorded in the Munsell system [@Munsell1947; @Simonson1993; @SoilSurveyManual2017] in the form of "hue value/chroma", the three components of this notation provide interpretive suggestions about iron oxides and oxidation state (hue and chroma) [@Schwertmann1993; @Scheinost1999], soil carbon (value) [@Wills2007; @Liles2013], as well as hints about the relative importance of catenary relationships [@Brown2004]. Several color-based metrics of soil development [@Buntley1965; @harden1982], rubification [@barron1986; @hurst1997], and melanization [@harden1982; @thompson1996] are implemented in the *aqp* package.


#### Color Conversion

The *aqp* package provides several interfaces for conversion between Munsell notation and sRGB or CIELAB color spaces, largely based on the 1943 Munsell renotation table [@centore2012]. Forward conversion from standard Munsell notation (e.g. 10YR 3/4) is performed via look-up table, derived from the renotation data and interpolated to include odd chroma values. The function `munsell2rgb()` performs a direct transformation to sRGB-encoded colors in hexadecimal (#5E4323), sRGB coordinates scaled to the interval of 0--1 ([0.36791, 0.26445, 0.13648]), or CIELAB coordinates ([30.807, 7.357, 24.027]) [@beaudette2013_AQP]. Inverse transformation from sRGB coordinates is performed by the `rgb2munsell()` function; approximated by nearest-neighbor search of the Munsell-sRGB look-up table using the CIE2000 color contrast metric [@farver_pkg2021]. All color space coordinates are referenced to the CIE standard illuminant D65, which is a close approximation to average midday sunlight in the northern hemisphere [@Marcus1998]. sRGB and CIELAB color spaces were selected to address two common uses of soil color: sRGB, due to the its use in most all computer graphics applications, and CIELAB for the convenient alignment of axes and common pigments in the soil environment [@Rossel_et_al_2006; @Liles2013]. 

Non-standard notation of Munsell colors (e.g. 10.6YR 3.3/5.5), as collected by digital colorimeter, can be converted to approximate sRGB coordinates using the `getClosestMunsellChip()` function. However, this approach uses rounding of value and chroma and snapping to the nearest standard hue (10YR). Exact conversion of non-standard Munsell notation can be performed using the *munsellinterpol* R package [@munsellinterpol_pkg2021].



#### Color Contrast

Color contrast (perceptual difference between two colors) within a soil sample is an important component of field-described redoximorphic features, concentrations, and mottles [@FieldBookv3]. Soil Survey products and wetland delineation protocols adopted by the National Cooperative Soil Survey (NCSS) currently use *contrast classes* (faint, distinct, and prominent) to describe color contrast, based on differences in Munsell hue, value, and chroma [@SSTN2]. The `colorContrast()` function in *aqp* computes differences in Munsell {hue, value, chroma}, soil color contrast class, and the CIE2000 color contrast metric [@Sharma2005] for pairs of colors specified in Munsell notation. The function is fully vectorized meaning that multiple comparisons can be generated without explicit looping (Table \@ref(tab:color-contrast-table)).

```{r color-contrast-table, echo = FALSE}
library(aqp)
cc <- colorContrast(c('10YR 3/3', '7.5YR 6/6', '2.5Y 2/2'), c('10YR 3/4', '5YR 4/6', '5G 4/8'))
names(cc) <- c('Color 1', 'Color 2', 'dH', 'dV', 'dC', 'dE00', 'CC')

knitr::kable(cc, digits = 1, align = 'c', caption = 'Output includes: change in hue (dH), change in value (dV), change in chroma (dC), CIE2000 color contrast (dE00), and NCSS soil color contrast class (CC).')
```

Tabular color contrast output can be convenient when used as an intermediate step in a more complex workflow, but can be difficult for non-specialists to interpret. A graphical representation of these data are created by the `colorContrastPlot()` function provided by *aqp*. For example, the differences between typical dry and moist soil colors for the Musick soil series are demonstrated in Figure \@ref(fig:contrast-class-figure). While exact replication of Munsell colors isn't possible on un-calibrated displays or printers, the sRGB approximation is sufficient to demonstrate relative differences in hue, value, and chroma.

```{r contrast-class-figure, fig.cap="Color contrast plot, comparing the moist and dry soil colors of the Musick soil series. CIE2000 color contrast values are printed below soil color contrast classes."}
knitr::include_graphics('figures/contrast-class-dE00.svg')
```

To further aid with the calculation and interpretation of color contrast, "color contrast charts" can be created with the `contrastChart()` function provided by *aqp*. These charts are based on a source color in Munsell notation (e.g. 7.5YR 4/3) and select pages of Munsell hue. Pair-wise metrics of color contrast are evaluated between all color "chips" and the source color (outlined in red), and arranged in a familiar format with Munsell chroma on the x-axis, value on the y-axis, and pages of hue split across panels. Soil color contrast class and CIE2000 values are printed below each color "chip" (Figure \@ref(fig:contrast-chart-figure)).

```{r contrast-chart-figure, fig.cap="Color contrast chart for 7.5YR 4/3, including reference 5YR, 7.5YR, and 10YR hue pages. Soil color contrast classes are abbreviated as F (faint), D (distinct), and P (prominent). Numbers are CIE2000 values."}
knitr::include_graphics('figures/contast-chart.svg')
```



### Soil Profile Sketches

Conceptual sketches of soil profiles that illustrate variation in morphology (e.g. horizon depths, horizon designations, color, texture, etc.) in relation to transect or catenary position are a pedologic staple. Either hand-drawn in field notes or carefully produced as part of a final soil survey manuscript, these sketches represent an important vehicle for communicating observation and context to technical and non-technical audiences alike. A data-driven approach to creating soil profile sketches was one the of the original motivations for the *aqp* R package [@beaudette2013_AQP]. Since 2010, the profile sketch authoring tools in *aqp* have progressed from basic layout of filled rectangles (profiles and horizons) to thematic coloring of horizons based on properties or classes, encoding of horizon boundary information, and handling of label collision to name a few examples.

```{r catena-sketch-figure, fig.cap="Illustration of an idealized hillslope catena for soil components from two adjacent mapunits within the Ozark Highlands.  SSURGO map unit soil components are approximately placed within a geomorphic hillslope context....."}
knitr::include_graphics('figures/cross-section.svg')
```

The `plotSPC()` function (*aqp* package) is the primary tool for creating soil profile sketches from `SoilProfileCollection` objects, using *R*'s "base graphics" system. Figure \@ref(fig:catena-sketch-figure) demonstrates several possible data sources, processing steps, and output generated from two calls to `plotSPC()`. Soil components (retrieved from the detailed Soil Survey via `fetchSDA()` as `SoilProfileCollection`) within map unit 2vxq8 occur on summit and shoulder hillslope positions while components in map unit 2vxq9 occur on backslope, footslope and toeslope positions. USDA soil texture classes used as colors to fill the soil profiles show the variation of textures within the catena. Labeling of horizon depths (vs. common depth axis), leader lines, and collision detection (common with thin horizons) are optional enhancements to the standard output, specified via function arguments [@aqp_pkg_manual]. Map unit names, component names (soil series), and parent material text are from the Soil Survey. Narrower profiles to the left of each component sketch represent data from the Official Series Descriptions via `fetchOSD()`. These data represent the typical morphology (horizon depths, designations, colors, etc.) for all soil series used in the US Soil Survey. Munsell colors (moist conditions) have been converted to sRGB coordinates using `munsell2rgb()`, and horizon boundary distinctness codes have been converted into vertical offsets using `hzDistinctnessCodeToOffset()`. The `plotSPC()` function can encode horizon distinctness offsets as diagonal horizon boundaries, where increasingly steeper angles represent the following sequence of boundary distinctness: "very abrupt", "abrupt", "clear", "gradual", "diffuse" [@FieldBookv3]. A visual explanation of the many arguments to `plotSPC()` is provided via `explainPlotSPC()` which shows the usage of ordering vectors, graphical offsets and scaling factors within the graphical space. Detailed examples of `plotSPC()` usage are available in the function documentation [@aqp_pkg_manual] and associated tutorials [@SPC_intro].  Future developments to `plotSPC()` will include conversion to the more advanced "grid" graphics system, pattern fills (e.g. geologic and stratigraphic symbols), and tighter integration with other plotting libraries such as *lattice* and *ggplot2*.



### Functional Horizon Aggregation

Soil scientists use a common language of horizon designation nomenclature to describe and articulate the observed differences in soil horizons within a soil profile. These basic notations and the act of 'naming' genetic horizons distills important information in the form of master horizons, characteristic subscripts, horizon and pedogenic sequences and parent material discontinuities.  Horizon designations convey a mental concept of the soil properties while allowing complexity and variability in how horizon designations are applied. Experienced soil scientists will generally apply horizon nomenclature similarly to the features they observe and identify in the process of describing a soil profile due to the rigid guidelines and definitions of their application.  Building on these similarities of horizon usage is a concept called, Generalized Horizon Labels (GHL) which seeks to unify a similar yet diverse set of soil profile horizonations for the purpose of aggregation, analysis, and summary operations[@Beaudette_DSMorph2016; @Roecker2016]. The process of applying GHL to a collection of soil profiles involves a series of micro-correlations made by a soil scientist to determine which horizons have similar soil morphology and properties to be grouped together for aggregation across horizons within a `SoilProfileCollection`. The aggregations into GHL provide a way to convey functional horizons for collections of soil profiles, such summary concepts can be extended to pedotransfer functions [@Wagenet1991].

The application of GHL to a `SoilProfileCollection` is driven by Regular Expression (REGEX) pattern matching. REGEX provides a powerful means of controlling and applying alpha-numeric string matching of horizonation sequences to 'collect' numerous varied horizon designation sequences into a set of unified horizon designations. A user developed set of REGEX rules are matched to an identified vector of horizonation by the `generalize.hz()` function in the *aqp* R package.  


```{r GHL-figure, fig.cap="A series of GHL applied to a collection of soil profiles of the Clarksville soil series from the Ozark Highlands, Missouri. Horizon labels within each soil profile show the original horizon designations while colors indicate assigned GHL."}
knitr::include_graphics('figures/sketch-demo.svg')
```


Features unique to the Clarksville soil series are argillic diagnostic horizons and parent material discontinuities as horizons transition along boundaries between cherty dolomite slope alluvium and colluvium over dolomite residuum. GHL can be used to identify these common features in a set of soils once applied (Figure \@ref(fig:GHL-figure)).

The `depthOf()` family of functions utilize REGEX pattern matching to specific horizon designation to calculate the depth of occurrence of that horizon designation within a collection of profiles. Given a set of Clarksville soil series profiles a `minDepthOf()` and `maxDepthOf()` operations provide additional utility to find either the top (shallowest) or bottom (deepest) depth to a matching horizon pattern. Once derived for a set of profiles they can be ordered according to the depth to the specified feature (e.g. '3Bt' as shown in Figure \@ref(fig:GHL-figure)). These functions provide convenience handling for missing values or when target patterns are not found within a profile. Results are returned as a numeric vector for single profiles or a `data.frame` of results with profile ID, horizon ID, top or bottom depths, horizon designation and pattern provided. Profile sketches in Figure \@ref(fig:GHL-figure) have been sorted by with values returned by the `depthOf()` function, first according to depth of "3Bt4" GHL and second according to depth of "2Bt3" GHL. 




### Change of Support


Soil data are typically collected either by genetic horizon (e.g. A, Bt, C [varying depths]), at regular depths (e.g. every 10cm), or from composite samples representing depth intervals (e.g. 0-25cm).  The structure of these depth intervals may be identical throughout the collection, but often vary from one profile to the next. To facilitate analysis throughout the profile collection, they may need to be modified and/or harmonized (i.e. change of support). A simple down-scaling of horizons (without interpolation) into a regular sequence of thinner depth slices, referred to as "slicing", is implemented in the `dice()` function (Figure \@ref(fig:slice-slab-GHL-figure)A). The `segment()` function offers another approach to restructuring horizon depths, using horizon-thickness weighted mean values for conversion to fixed depth intervals (e.g. 0-25cm). This is is a common step in the thematic mapping of soil property data. A more complex change of horizon depths can be achieved using constrained interpolation. This method, popular for applications such as digital soil mapping and requiring harmonized horizon depths, was proposed by @bishop1999modelling, and uses mass-preserving splines. In *aqp*, this type of down-scaling is performed with the `spc2mpspline()` function--a convenient interface to the *mpspline2* package [@mpspline2_pkg2020] suitable for `SoilProfileCollection` objects.

<!-- adapted from 2015 DSMorph ideas / figure -->
```{r slice-slab-GHL-figure, fig.cap="Change of support demonstration using 25 pedons correlated to the Clarksville soil series. Generalized horizon labels (GHL) have been resampled to 1cm slices (A) via `dice()`, aggregated across slices to GHL proportions (B) via `slab()`, and modeled via proportional-odds logistic regression (C)."}
knitr::include_graphics('figures/slice-slab-GHL-both.svg')
```

A change of support operation performed over all profiles in a collection results in a statistical summary of that collection. The `slab()` function performs this kind of operation by "slicing" horizon data (continuous or categorical) into 1cm-thick depth intervals (Figure \@ref(fig:slice-slab-GHL-figure)A), then aggregating over those depth intervals [@harradine1963; @beaudette2013_AQP]. Continuous values are reduced to select percentiles (or any user-defined function) per slice, and categorical values are reduced to proportions (Figure \@ref(fig:slice-slab-GHL-figure)B) [@beaudette2013_terrain]. Alternatively, sliced categorical data such as generalize horizon labels (GHL) can be aggregated using statistical models for ordinal data--such as the proportional-odds logistic regression model (Figure \@ref(fig:slice-slab-GHL-figure)C) [@Beaudette_DSMorph2016].

